Local simple connectedness of boundaries of hyperbolic groups

### 2020-04-16 paper.tex
Added todos from comments

\todo{Proof-read section 3 onwards}

@@

If a group acts geometrically on a proper hyperbolic space then, following
Gromov~\cite{gromov87}, we say that the group is hyperbolic, and quasi-isometry
invariants of the space can be thought of as invariants of the group. One
particularly powerful quasi-isometry invariant that can be used in
this way is the Gromov boundary of the space. This is the set of geodesic rays
in the space eminating from a fixed base point, where rays are defined to be
equivalent if they remain a bounded distance apart. In this way one can define
the boundary $\boundary G$ of a hyperbolic group $G$.

@@

In~\cite{gromov87}, Gromov showed that the boundary of a hyperbolic group is a
compact metric space with finite topological dimension. Beyond these basic
properties, and excepting the case when $G$ is the fundamental group of a
negatively curved smooth manifold, in which case $\boundary G$ is a sphere,
boundaries of hyperbolic groups tend to be wild, fractal-like objects.
In~\cite{kapovichkleiner00}, Kapovich and Kleiner ask the following question.

@@

On the one hand, finding concrete examples of spaces arising as boundaries has
proved difficult, but on the other, relatively few restrictions on the topology
of the boundary are known. One restriction of particular importance is a
theorem of Bestvina and Mess~\cite{bestvinamess91}, which states that if
$\boundary G$ is connected and is not disconnected by the removal of any single
point then $\boundary G$ is locally path connected. Combined with a result of
Bowditch~\cite{bowditch98b} and Swarup~\cite{swarup96}, this implies that
$\boundary G$ is locally path connected whenever it is connected.

@@

@@

The paper is structured as follows. In Section~\ref{sec:spheres} we use the
Rips complex of the group $G$ to build a sequence $(K_n)_{n \geq 0}$ of
simplicial complexes that give successively better approximations to the
boundary $\boundary G$. These simplicial complexes are spheres in the Rips
complex. There exist natural simplicial maps $p^n_m\from K_n\to K_m$ for $n -
m$ large enough, and there exist natural continuous maps
$p^\infty_n\from\boundary G \to K_n$ for all $n$.  The remainder of
Section~\ref{sec:spheres} will be spent proving quantitative statements about
these maps.

@@

In Section~\ref{sec:double_dagger} we begin by recalling the so-called
double-dagger condition $\ddag_M$ of Bestvina and Mess. This condition is
guaranteed to hold if $\boundary G$ does not contain a cut point, and implies
that $\boundary G$ is locally path connected. We then modify the condition to
allow for easier generalisation, and introduce a new condition $\S_M$, which we
will see to be related the local simple connectedness of $\boundary G$. Using
the compactness results from Section~\ref{sec:horoball} we show that this
condition is satisfied whenever we have that $\boundary G - \set\xi$ is simply
connected for any $\xi$ in $\boundary G$.

Then, in Section~\ref{sec:local_simple_connectedness} we show that our new
condition $\S_M$ implies the local simple connectedness of $\boundary G$. We
begin by using $\S_M$ to construct a sequence of maps $(i^n_\infty)_n$ from the
2-skeleton of $K_n$ to $\boundary G$. These maps can be thought of as
approximate sections of $p^\infty_n$, in that $i^n_\infty\composed p^\infty_n$
converges uniformly to the identity map $\boundary G \to \boundary G$ as $n$
tends to $\infty$. By controlling the diameter of the image under $i^n_\infty$
of a simplex in $K_n$ we are able to prove the local simple connectedness of
$\boundary G$.

@@

\begin{remark}\label{rem:dist_vs_newdist}
  The extended ``metric'' $\newdist$ is not a metric, since it only satisifes
  the triangle inequality up to a bounded additive error, but this will not
  cause a problem.

@@

\begin{definition}
  \todo{Extend to the boundary}
  Given points $x$, $y$ and $z$ in $K$, define the \emph{Gromov product
  $\gromov{x}{y}{z}$ of $x$, $y$ and $z$} as follows.
  \begin{equation*}
    \gromov{x}{y}{z} = \frac{1}{2}\left(\newdist(x,y) + \newdist(x,z) -
          \newdist(y,z)\right)
  \end{equation*}
\end{definition}

@@

\begin{proposition}\label{prop:i_in_dimension_1}
  Suppose that $G$ satisfies $\ddag'_{D + 4\delta+2}$. Then there exist $L$ and
  $n_0$ such that for $n\geq n_0$ there is a map $i^n_{n+1}\from \Skel_1 K_n
  \to \Skel_1 K_{n+1}$ satisfying the following conditions:
  \begin{enumerate}
    \item for each $n$, $i^n_{n+1}$ is a simplicial map on the subdivision
      $\sd_L\Skel_1 K_n$,
    \item for any vertex $x$ of $K_n$, $\dist(x, i^n_{n+1}(x)) \leq 2\delta+1$,
      and
    \item for each $n$, the map $i^n_{n+1}$ is a simplicial approximation a map
      that factors through $p^\infty_{n+1}\from \boundary G \to K_{n+1}$.
  \end{enumerate}
\end{proposition}

@@

\begin{proposition}\label{prop:i_in_dimension_2}
  There are constants $n_0$ and $L$ such that for each $n\geq n_0$ there is map
  $i^n_{n+1}\from\Skel_2 K_n \to \Skel_2 K_{n+1}$ with the following properties.
  \begin{enumerate}
    \item For each $n$, $i^n_{n+1}$ is an affine map on the subdivision
      $\sd_L\Skel_2 K_n$.
    \item For any vertex $x$ of $K_n$, $\dist(x, i^n_{n+1}(x)) \leq 2\delta+1$.
  \end{enumerate}
\end{proposition}

@@

\begin{lemma}\label{lem:convergestoidentity}
  There is a constant $C$ such that the distance from $i^m_\infty \composed 
  p^\infty_m$ to the identity map $\boundary G\to\boundary G$ is at most 
  $Ca^{-m}$ for each $m \geq n_0$.
\end{lemma}

@@

  Let $U$ be a neighbourhood of $\zeta_+$ such that the closure of $U$ does not
  contain $\zeta_-$ and let $V$ be a neighbourhood of $\zeta_+$ contained in
  $U$ such that any map from the circle to $V$ is nullhomotopic in $U$. Then by
  the properties of the sequence $(g_n)$ described above, there exists an
  element $g_n$ of the sequence such that $g_n\cdot\xi \notin U$ and
  $g_n\composed\gamma$ has image in $V$. Then $g_n\composed\gamma$ admits a
  contracting homotopy with image in $U$, and the image of this homotopy under
  $g_n^{-1}$ is a contracting homotopy of $\gamma$ with image contaed in
  $\boundary G - \set\xi$.
\end{proof}

@@

It is also natural to ask whether the theorem of this paper generalises to
higher dimensions. The difficulty is in generalising the proof of
Proposion~\ref{prop:sufficient_condition}: given a sequence of maps from $S^0$
to $\boundary G$ we pass to a uniformly convergent subsequence, but this is not
possible for higher dimensional spheres without controlling the modulus of
continuity of the maps.

>>>>

\begin{abstract}
  In this paper we prove a theorem describing the local topology of the
  boundary of a hyperbolic group in terms of its global topology: the boundary
  is locally simply connected if and only if the complement of any point in the
  boundary is simply connected. This generalises a theorem of Bestvina and
  Mess~\cite{bestvinamess91}.
\end{abstract}

@@

If a group acts geometrically on a proper hyperbolic metric space then,
following Gromov~\cite{gromov87}, we say that the group is hyperbolic, and
quasi-isometry invariants of the space can be thought of as invariants of the
group. One particularly powerful quasi-isometry invariant that can be used in
this way is the Gromov boundary of the space. This a topological space defined
as the set of equivalence classes of geodesic rays in the space emanating from
a fixed base point, where rays are defined to be equivalent if they remain a
bounded distance apart. In this way one can define the boundary $\boundary G$
of a hyperbolic group $G$.

@@

In~\cite{gromov87}, Gromov showed that the boundary of a hyperbolic group is a
compact metric space with finite topological dimension. Beyond these basic
properties, and excepting the case when $G$ is the fundamental group of a
negatively curved smooth closed manifold, in which case $\boundary G$ is a
sphere, boundaries of hyperbolic groups tend to be wild, fractal-like objects.
In~\cite{kapovichkleiner00}, Kapovich and Kleiner ask the following question.

\todo{Survey of examples}

@@

On the one hand, finding concrete examples of spaces arising as boundaries has
proved difficult, but on the other, relatively few restrictions on the topology
of the boundary are known. One restriction of particular importance is a
theorem of Bestvina and Mess~\cite{bestvinamess91}, which states that if
$\boundary G$ is connected and cannot be disconnected by the removal of any
single point then $\boundary G$ is locally path connected. Combined with a
result of Bowditch~\cite{bowditch98b} and Swarup~\cite{swarup96}, this implies
that $\boundary G$ is locally path connected whenever it is connected.

@@

\todo{Example of space ruled out by theorem}

@@

The paper is structured as follows. In Section~\ref{sec:spheres} we use the
Rips complex of the group $G$ to build a sequence $(K_n)_{n \geq 0}$ of
simplicial complexes that give successively better approximations to the
boundary $\boundary G$. These simplicial complexes are spheres in the Rips
complex. There exist natural simplicial maps $p^n_m\from K_n\to K_m$ for $n \gg
m$, and there exist natural continuous maps $p^\infty_n\from\boundary G \to
K_n$ for all $n$.  The remainder of Section~\ref{sec:spheres} will be spent
proving quantitative statements about these maps.

@@

In Section~\ref{sec:double_dagger} we begin by recalling the so-called
double-dagger condition $\ddag_M$ of Bestvina and Mess. This condition is
guaranteed to hold if $\boundary G$ does not contain a cut point, and implies
that $\boundary G$ is locally path connected. We then modify the condition to
allow for easier generalisation, and introduce a new condition $\S_M$, which we
will see to be related the local simple connectedness of $\boundary G$. Using
the compactness results from Section~\ref{sec:horoball} we show that this
condition is satisfied if $\boundary G - \set\xi$ is simply connected for any
$\xi$ in $\boundary G$.

\todo{Maybe rephrase?}
Then, in Section~\ref{sec:local_simple_connectedness} we show that our new
condition $\S_M$ implies the local simple connectedness of $\boundary G$. We
begin by using $\S_M$ to construct a sequence of maps $(i^n_\infty)$ from the
2-skeleton of $K_n$ to $\boundary G$. These maps can be thought of as
approximate sections of $p^\infty_n$, in the following sense. For each $n$, let
$r_n\from K_n \to \Skel_2 K_n$ be a (not necessarily continuous) map sending
points to nearby points in the 2-skeleton. Then $i^n_\infty\composed r_n
\composed p^\infty_n$ converges uniformly to the identity map $\boundary G \to
\boundary G$ as $n$ tends to $\infty$. By controlling the diameter of the image
under $i^n_\infty$ of a simplex in $K_n$ we are able to prove that $\boundary
G$ is locally simply connected.

@@

\begin{remark}\label{rem:dist_vs_newdist}
  The extended ``metric'' $\newdist$ is not a metric, since it only satisfies
  the triangle inequality up to a bounded additive error, but this will not
  cause a problem.

@@

\begin{definition}
  Let $x \in K$ and let $y$ and $z$ be points in $K \union \boundary G$. Let
  $x^\star$ be a vertex of the minimal simplex of $K$ containing $x$. If $y$
  (respectively $z$) is in $\boundary G$ let $y^\star = y$ (respectively
  $z^\star = z$); otherwise let $y^\star$ (respectively $z^\star$) be a vertex
  of the minimal simplex of $K$ containing $y$ (respectively $z$.)
  Then define the \emph{Gromov product $\gromov{x}{y}{z}$ of $x$, $y$ and $z$}
  as $\gromov{x}{y}{z} = \gromov{x^\star}{y^\star}{z^\star}$.
\end{definition}

@@

\begin{proposition}\label{prop:i_in_dimension_1}
  Suppose that $G$ satisfies $\ddag'_{D + 4\delta+2}$. Then there exist $L$ and
  $n_0$ such that for $n\geq n_0$ there is a map $i^n_{n+1}\from \Skel_1 K_n
  \to \Skel_1 K_{n+1}$ satisfying the following conditions:
  \begin{enumerate}
    \item for each $n$, $i^n_{n+1}$ is a simplicial map on the subdivision
      $\sd^L\Skel_1 K_n$,
    \item for any vertex $x$ of $K_n$, $\dist(x, i^n_{n+1}(x)) \leq 2\delta+1$,
      and
    \item for each $n$, the map $i^n_{n+1}$ is a simplicial approximation a map
      that factors through $p^\infty_{n+1}\from \boundary G \to K_{n+1}$.
  \end{enumerate}
  (When $X$ is a simplicial complex, we denote by $\sd^n X$ the complex
  obtained from $X$ by barycentrically subdividing $n$ times.)
\end{proposition}

@@

\begin{proposition}\label{prop:i_in_dimension_2}
  There are constants $n_0$ and $L$ such that for each $n\geq n_0$ there is map
  $i^n_{n+1}\from\Skel_2 K_n \to \Skel_2 K_{n+1}$ with the following properties.
  \begin{enumerate}
    \item For each $n$, $i^n_{n+1}$ is an affine map on the subdivision
      $\sd^L\Skel_2 K_n$.
    \item For any vertex $x$ of $K_n$, $\dist(x, i^n_{n+1}(x)) \leq 2\delta+1$.
  \end{enumerate}
\end{proposition}

@@

\todo{Doesn't quite make sense.}
\begin{lemma}\label{lem:convergestoidentity}
  There is a constant $C$ such that the distance from $i^m_\infty \composed 
  p^\infty_m$ to the identity map $\boundary G\to\boundary G$ is at most 
  $Ca^{-m}$ for each $m \geq n_0$.
\end{lemma}

@@

  Let $U$ be a neighbourhood of $\zeta_+$ such that the closure of $U$ does not
  contain $\zeta_-$ and let $V$ be a neighbourhood of $\zeta_+$ contained in
  $U$ such that any map from the circle to $V$ is null-homotopic in $U$. Then by
  the properties of the sequence $(g_n)$ described above, there exists an
  element $g_n$ of the sequence such that $g_n\cdot\xi \notin U$ and
  $g_n\composed\gamma$ has image in $V$. Then $g_n\composed\gamma$ admits a
  contracting homotopy with image in $U$, and the image of this homotopy under
  $g_n^{-1}$ is a contracting homotopy of $\gamma$ with image contained in
  $\boundary G - \set\xi$.
\end{proof}

@@

It is also natural to ask whether the theorem of this paper generalises to
higher dimensions. The difficulty is in generalising the proof of
Proposition~\ref{prop:sufficient_condition}: given a sequence of maps from $S^0$
to $\boundary G$ we pass to a uniformly convergent subsequence, but this is not
possible for higher dimensional spheres without controlling the modulus of
continuity of the maps.
